---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
---

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# Install packages
paket <- function(pak){
  new_pak <- pak[!(pak %in% rownames(installed.packages()))]
  if (length(new_pak)) 
    install.packages(new_pak, dependencies = TRUE,repos="https://cloud.r-project.org/")
  sapply(pak, library, character.only = TRUE)
}

listOfPackages <- c("tidyverse", "RColorBrewer", "kableExtra")
paket(listOfPackages)
```


```{r}
# options(repr.plot.width = 18, repr.plot.height = 9)
theme_plots <- theme_bw() +
  theme(strip.text = element_text(size = 5),
        axis.text.x = element_text(size = 9), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11, face = "italic")) 
theme_set(theme_plots)
```


# Inclusion criteria
The inclusion criteria for this study is:
- inpatients
- Sars-Cov-2 PCR positive or negative (no filter by covid-19 condition, just PCR test done at BCH)
- filtered by age (11-17)
- 1st January 2019 - Up to date
- The ICD-10 code from the list as diagnosis during hospitalization

# Variables that need to be checked/modified by each site 
Change the values of the next variables to your specific site:
- "myinputFiles": the specific path where your phase 2.2 data files are located. 
- "obfuscation": determine the obfuscation threshold (FALSE if no obfuscation, the numeric value of the obfuscation threshold if any)
- "raceAvailable": set up as TRUE or FALSE depending on if the variable is collected or not at your site
- "dateFormat": specify the format of the date in your site (e.g, 03-AUG-20, the format would be "%d-%b-%y")

```{r message=FALSE, warning=FALSE}
myinputFiles <- "/4ceData/Input/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
```

# Data input
We will use as input the 2.2 data files. Specifically: 
- localpatientsummary
- localpatientobservation
- localpatientclinicalcourse

For those sites recording the race, we will also read an additional file:
- localPatientRace

## Read the patient level data input files.
We will read the 2.2. patient level data files. 

```{r message=FALSE, warning=FALSE}
### Read the CSV input files
source("R/readInputFiles.R")

files <- readInputFiles( path      = myinputFiles, 
                         separator = ",",
                         skip      = 0, 
                         verbose   = FALSE )
  
### Extract the patient summary and observation information. 
demo_raw <- files[[1]]
obs_raw <- files[[2]]
clinical_raw <- files[[3]]

### Read the file containing race information for those sites recording this variable
if( raceAvailable == TRUE ){
  race_raw <- read.delim(paste0( myinputFiles, "/LocalPatientRace.csv"), sep = ",", skip = 0)
}
```

## Read the ICD10 codes file
We also need to read the file containing the ICD10 psychiatric related codes, located in the public-data folder of the GitHub repository. 

```{r suppData, message=FALSE}
icdCodes <- read.csv("public-data/psychiatric_ICD-10_codes_adjusted.csv", header = TRUE, colClasses = "character")

#get only the code 
icdCodes$concept_code <- sapply(strsplit( icdCodes$ICD10_Code, "[:]"), '[', 2)

# remove potential duplicates and select the columns of interest
icdCodes <- icdCodes %>% 
  dplyr::group_by( concept_code ) %>%
  dplyr::summarise( disorder_group = Mental_Health_Disorder_Group[1], 
                       description = Description[1]) %>%
  dplyr::select( disorder_group, concept_code, description )
```


# Data transformation
From the whole dataset we will select only those patients that have been diagnosed with at least one of the mental health ICD10 codes. 
For these patients we will generate a table that will allow us to perform all the analysis required for this study. 
This transformed table will contain 9 columns:
- patient_num: patient internal identifier
- sex (male, female)
- age (in years)
- concept_type (DIAG-ICD10)
- concept_code (ICD-10 full code, e.g, R44.1)
- calendar_date (day-month-year, 30-SEP-20)
- hospitalization status (1, 0)
- n_hospitalization (the hospitalization number)
- len_hospitalization (length of each hospitalization)

The last two columns are not available in the input data files. We will generate those counts using the information available in the clinical raw file.  

## Adding columns with hospitalization number and hospitalization length to the clinical raw file
```{r}
count_sequences_hospitalisation <- function(df, ...) {
  count_sequences <- rle(df$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  df[df$in_hospital == 1, "n_hospitalisation"] <- sequences
  df[df$in_hospital == 1, "len_hospitalisation"] <- sequences_len
  return(df)
}
clinical_raw <- clinical_raw %>% group_by(patient_num) %>% 
  group_modify(count_sequences_hospitalisation)
```

## Creating the table with all the data we need after filtering by inclusion criteria
Now that we have all the information need it, we create our input table:
- selecting only patients with mental diagnosis during hospitalization
- add the demographic and hospitalization related data 
- select only patients from 11 to 17 years old. 
- add the concept description and the disorder group

We will also transform the date, and break by week, adding 2 additional columns, "weeks" and "year". 
Finally, based on a cut-off date, we will determine if the diagnosis was done in early or late pandemic. 
For those sites recording the race of the patients, we also add the information to the main table. 

### Check that we have patients with the ICD10 mental codes
```{r}
codeLength <- unique( obs_raw %>%
  dplyr::filter( concept_type == "DIAG-ICD10") %>%
  dplyr::mutate( concept_code = as.character( concept_code ), 
                 concept_length = nchar( concept_code ), 
                 concept_check = ifelse( concept_length == 4 | concept_length > 8, "Check", "OK")) %>%
  dplyr::select( concept_code, concept_length, concept_check) )

summary(as.factor( codeLength$concept_check ) )

### check regular expression [A-TV-Z][0-9][0-9AB]\.?[0-9A-TV-Z]{0,4}

checkCodes <- icdCodes %>%
  dplyr::filter( concept_code %in% obs_raw$concept_code ) %>%
  dplyr::select( concept_code)

print( paste0("There are ", length(checkCodes$concept_code), " ICD10 mental codes out of the total ",length(unique(icdCodes$concept_code)), " in the dataset"))
```


### Generate the table
```{r message=FALSE, warning=FALSE}
cut_off_date <- "2021-04-01"

input_data <- obs_raw %>% 
  full_join( clinical_raw,  by = c("patient_num", "days_since_admission")) %>%
  full_join( demo_raw, by = "patient_num") %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ), weeks = cut( date, breaks= "week"), year = format( date, "%Y"), 
                 period = ifelse( date < cut_off_date, "early_pandemic","late_pandemic")) %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 in_hospital == 1, 
                 concept_code %in% icdCodes$concept_code, 
                 age < 18 & age > 10,
                 date >= "2019-01-01") %>%
  left_join( icdCodes, by = "concept_code") %>%
  dplyr::select( patient_num, sex, age, concept_type, concept_code, calendar_date, in_hospital, n_hospitalisation, len_hospitalisation, 
                 date, weeks, year, period, disorder_group, description )

if( raceAvailable == TRUE ){
  input_data <- input_data %>%
    left_join( race_raw, by = "patient_num" )
}
```


# Data analysis 

## Patients with a disorder group previous to the pandemic vs patients that present the disorder group the first time during the pandemic
We estimate the number of patients that a disorder group:
- only before the pandemic
- only after the pandemic
- both, before and after the pandemic 

The output is a barplot showing in the x-axis the number of patients, and in the y-axis the distinct disorder groups. Each disorder group has 3 bars associated based on the three previous options. 

```{r}
summary_by_group_and_period <- input_data %>% 
  dplyr::group_by( patient_num, disorder_group ) %>%
  dplyr::mutate( 
    when = ifelse( "early_pandemic" %in% period & "late_pandemic" %in% period,
                   paste("Before and After", cut_off_date),
                   ifelse( "early_pandemic" %in% period & ! "late_pandemic" %in% period,
                           paste("Only Before", cut_off_date),
                           paste("Only After", cut_off_date)
                   )
    ) %>% factor(levels = c(
      paste("Only Before", cut_off_date), 
      paste("Before and After", cut_off_date),
      paste("Only After", cut_off_date)
    ))
  ) %>%
  dplyr::select( patient_num, when)

summary_by_group_and_period %>%
  dplyr::group_by( disorder_group, when ) %>%
  dplyr::summarise( count= n_distinct( patient_num ) ) %>%
  ggplot(aes(x = as.factor( disorder_group), y = count, fill = when)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_brewer(palette = "Blues", name = "Period") +
  labs(y = "Unique patients count",
       x = "Disorder group",
       title = "Patients with at least one visit for a disorder group", 
       subtitle = "According to visit period", 
       legend = "Period")+
  theme(legend.position = "bottom", 
        legend.direction = "vertical",
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0)) + 
  coord_flip() +
  theme_bw()
  # scale_fill_grey()

ggsave("output/disorder_before_after_pandemic.png", width = 18, height = 9, units="cm")
```

## Patients count per week
Here we estimate the number of patients hospitalized per week. 
The output is a barplot showing in the x-axis the week, and in the y-axis the patient counts. Additionally a temporal trend line is added on top of the bars.
```{r}
#estimate the number of patients per week
input_data[, c("patient_num", "weeks")] %>%
  unique() %>%
  group_by(weeks) %>%
  tally() %>%
  mutate(n = ifelse( n > obfuscation, n, obfuscation * 0.5)) %>%
  ggplot(aes(x = as.Date(weeks), y = n, fill = "Patients count")) +
  geom_bar(stat="identity")+
  labs(y = "Counts",
       x = "Week",
       title = "Weekly patient counts with mental health related ICD codes",
       subtitle = "BCH - Inpatient diagnoses, X-Y age group") + 
  theme(legend.position = "none") + 
  scale_fill_brewer(palette = "Dark2") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_weekly_count.png", width = 18, height = 9, units="cm")
```

We replicate the previous graphic aggregating by:
- Sex 
- Age group

### Patients count per week aggregated by sex
```{r}
input_data[, c("patient_num", "sex","weeks")] %>%
  unique() %>%
  group_by(weeks, sex) %>%
  tally() %>%
  mutate(n = ifelse( n > obfuscation, n, obfuscation * 0.5)) %>%
  ggplot(aes(x = as.Date(weeks), y = n, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  labs(y = "Counts",
       x = "Week",
       title = "Weekly patient counts with mental health related ICD codes aggregated by sex",
       subtitle = "BCH - Inpatient diagnoses, X-Y age group") + 
  theme(legend.position = "bottom") + 
  #scale_fill_grey() +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```

### Patients count per week aggregated by age group 
This graphic will only make sense if we include a wide range of age. 
Right now this comment is not run since we are only including adolescents. 

```{r, eval=FALSE}
#estimate the number of patients per week by age group
input_data[, c("patient_num", "age","weeks")] %>%
  unique() %>%
  mutate( age_group = ifelse( age < 10, "children", "adolescent")) %>%
  group_by(weeks, age_group) %>%
  tally() %>%
  mutate(n = ifelse( n > obfuscation, n, obfuscation * 0.5))  %>%
    ggplot(aes(x = as.Date(weeks), y = n, fill = age_group)) +
    geom_bar(stat="identity", position = "dodge")+
    labs(y = "Counts",
         x = "Week",
         title = "Weekly patient counts with mental health related ICD codes aggregated by age group",
         subtitle = "BCH - Inpatient diagnoses, X-Y age group") + 
    theme(legend.position = "bottom") + 
    scale_fill_grey() +
    geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)

```

## Patients counts per week rouping by ICD subgroup category 
Here we estimate the number of patients hospitalized per week but we break it based on the distinct disorder groups. 
The output is a barplot showing in the x-axis the week, and in the y-axis the patient counts. Additionally a temporal trend line is added on top of the bars.
Each barplot represent a disorder group.

```{r}
input_data[input_data$in_hospital == 1,c("patient_num", "weeks", "disorder_group")] %>%
  unique() %>%
  group_by(weeks, disorder_group) %>% tally() %>%
  mutate(n = ifelse( n > obfuscation, n, obfuscation * 0.5)) %>%
  ggplot(aes(x = as.Date(weeks), y= n, fill = "", group = disorder_group )) +
  geom_bar(stat = "identity")+
  labs(x = "Week",
       y = "Counts",
       title = "Patient counts with mental health related ICD codes", 
       subtitle = "Grouped by ICD subcategories") + 
  scale_fill_manual(values=c('#00BFC4'))+
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 5),
        axis.text.x = element_text(size = 9), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none")
ggsave("output/categorical_counts_inpatient_teens.png", width = 18, height = 9, units="cm")

```

# Summary table 1
Here we will estimate the counts to populate table 1, with the counts for early and late pandemic, and a statistical test comparing both. 
This table one will include:
- sex
- age
- race (when available)
- hospitalization length
- diagnosis codes 
- diagnosis groups

```{r}
t.test2 <- function(m1, m2, s1, s2, n1, n2) {
  # t.test computed from sample statistics, ie using mean, standard deviation and count value of the samples
  # useful in case we can only retrieve aggregated stats from the sites
  # ref: https://stats.stackexchange.com/questions/30394/how-to-perform-two-sample-t-tests-in-r-by-inputting-sample-statistics-rather-tha 
  # and https://en.wikipedia.org/wiki/Student%27s_t-test?oldformat=true#Equal_or_unequal_sample_sizes,_unequal_variances_(sX1_%3E_2sX2_or_sX2_%3E_2sX1)
  se <- sqrt( (s1^2/n1) + (s2^2/n2) )
  # welch-satterthwaite df
  df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
  t <- (m1-m2)/se 
  dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))    
  names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
  return(dat)
}
```

## Age
```{r}
age_stats <- input_data %>%
  dplyr::distinct(patient_num, period, age) %>% 
  dplyr::group_by( period ) %>% 
  dplyr::summarise( mean = mean( age ), 
                    sd = sd( age ), 
                    count = n()) %>% 
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::ungroup() %>% 
  dplyr::bind_rows(
    dplyr::summarise(input_data, period="Total", mean=mean(age), sd=sd(age), count = n())
  )
age_test <- t.test2(m1 = age_stats[[1, "mean"]], 
                    m2 = age_stats[[2, "mean"]], 
                    s1 = age_stats[[1, "sd"]], 
                    s2 = age_stats[[2, "sd"]], 
                    n1 = age_stats[[1, "count"]], 
                    n2 = age_stats[[2, "count"]]) %>% 
  `[[`("p-value")
```

## Sex
```{r}
sex <- input_data %>% 
  dplyr::group_by( sex ) %>%
  dplyr::summarise( count = n_distinct(patient_num) ) %>%
  dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5), 
                 total = sum( count ), percentage = round( count/total*100, 2)  )

sex_by_period <- input_data %>% 
  group_by(period, sex ) %>% 
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, .add=TRUE) %>% 
  mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
          percentage=round(count/sum( count )*100, 2))
  

sex_test <- sex_by_period %>% 
  pivot_wider(names_from = sex, values_from = c(count, percentage)) %>% 
  ungroup() %>% 
  select(count_female, count_male) %>% 
  chisq.test() %>% 
  `[[`("p.value")
```

## Race
```{r}
if( raceAvailable == TRUE ){
  race <- input_data %>%
    dplyr::group_by( race_4ce ) %>%
    dplyr::summarise( count = n_distinct(patient_num) ) %>%
    dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5), 
                   total = sum( count ), percentage = round( count/total*100, 2)  )
  
  race_by_period <- input_data %>%
    group_by(period, race_4ce ) %>% 
    summarise(count = n_distinct(patient_num)) %>%
    group_by(period, .add=TRUE) %>% 
    mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
            percentage=round(count/sum( count )*100, 2))
    
  
summary_race_test <- race_by_period %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  select(late_pandemic, early_pandemic) %>% 
  tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")
}
```

## Diagnosis

### By ICD code
```{r}
summary_diagnosis <- input_data %>%
  dplyr::distinct( patient_num, description, period, weeks, concept_code ) %>% 
  dplyr::group_by( concept_code, description, period ) %>%
  dplyr::count( patient_num ) %>% 
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc(count) ) %>% 
  dplyr::ungroup()

summary_diagnosis_test <- summary_diagnosis %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  select(late_pandemic, early_pandemic) %>% 
  tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")

```

### By disorder group
```{r}
summary_disorder_group <- input_data %>%
  dplyr::group_by( disorder_group ) %>%
  dplyr::count( patient_num ) %>% 
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc( count ))

summary_disorder_group_by_period <- input_data %>% 
  dplyr::distinct( patient_num, disorder_group, period, weeks ) %>% 
  dplyr::group_by( period, disorder_group ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, add=TRUE) %>%
  mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5), 
         percentage=round(count/sum( count )*100, 2)) 

summary_disorder_group_by_period_test <- summary_disorder_group_by_period %>% 
  pivot_wider(id_cols = disorder_group, names_from = period, values_from = c(count, percentage)) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, replace = 0))) %>% 
  select(count_early_pandemic, count_late_pandemic) %>% 
  chisq.test() %>% 
  `[[`("p.value")

```


## Hospitalization

### Length of hospitalization 
```{r}
## Calculating also mean and sd in case we can only retrieve aggregated data from the sites
hospit_length <- input_data %>%
  dplyr::distinct( patient_num, n_hospitalisation, len_hospitalisation, period ) %>%
  dplyr::group_by( period ) %>% 
  dplyr::summarise( n_hospitalisations = n(),
                    length_hospt_median = median(len_hospitalisation), 
                    IQR = IQR( len_hospitalisation), 
                    length_hospt_mean = round(mean(len_hospitalisation), 2),
                    length_hospt_sd = round(sd(len_hospitalisation), 2))

hospit_length_mean_test <- t.test2(
  m1 = hospit_length[[1, "length_hospt_mean"]],
  m2 = hospit_length[[2, "length_hospt_mean"]],
  s1 = hospit_length[[1, "length_hospt_sd"]],
  s2 = hospit_length[[2, "length_hospt_sd"]],
  n1 = hospit_length[[1, "n_hospitalisations"]],
  n2 = hospit_length[[2, "n_hospitalisations"]]
)

```

### Repeat hospitalizations by patient
```{r}
hospitalisation_count <- input_data %>% 
  dplyr::distinct( patient_num, period, n_hospitalisation ) %>% 
  dplyr::group_by( period, patient_num ) %>%
  dplyr::filter(n_hospitalisation == max(n_hospitalisation)) %>%
  dplyr::group_by( period, add = FALSE) %>% 
  dplyr::summarise( mean_count_hospitalizations = mean(n_hospitalisation), 
                    max_count_hospitalizations = max(n_hospitalisation),
                    median_count_hospitalizations = median(n_hospitalisation), 
                    sd_hospitalisation_count = sd(n_hospitalisation),
                    IQR_hospitalisation_count = IQR( n_hospitalisation ))

repeated_hospitalisation <- input_data %>% 
  dplyr::distinct( patient_num, period, n_hospitalisation ) %>% 
  dplyr::group_by(patient_num, period) %>% 
  dplyr::filter(n_hospitalisation == max(n_hospitalisation)) %>% 
  dplyr::mutate(repeated_hospitalisation = if_else(n_hospitalisation > 1, "repeated", "non_repeated")) %>% 
  dplyr::group_by( period, add = FALSE ) %>% 
  dplyr::count( repeated_hospitalisation ) %>% 
  tidyr::pivot_wider(names_from = repeated_hospitalisation, values_from = n)
repeated_hospitalisation
repeated_hospitalisation_test <- chisq.test(repeated_hospitalisation[, c("non_repeated", "repeated")]) %>% 
  `[[`("p.value")
```


## Pulling together Table 1 statistics
```{r}
format_counts <- function(count, percentage) {
  glue::glue("{count} ({percentage} %)", count = count, percentage = percentage)
}

sex_table1 <- sex_by_period %>% 
  bind_rows(
    sex_by_period %>%
      group_by(period) %>% 
      dplyr::summarise(sex = "Total", count = sum(count), percentage = sum(percentage))
  ) %>% 
  mutate(result = format_counts(count, percentage)) %>% 
  select(-c(count, percentage)) %>% 
  pivot_wider( names_from = period, values_from = result ) %>% 
  rename(name = sex)

if( raceAvailable == TRUE ){
  race_table1 <- race_by_period %>% 
  bind_rows(
    race_by_period %>%
      group_by(period) %>% 
      dplyr::summarise(race_4ce = "Total", count = sum(count), percentage = sum(percentage))
  ) %>% 
  mutate(result = format_counts(count, percentage)) %>% 
  select(-c(count, percentage)) %>% 
  pivot_wider( names_from = period, values_from = result ) %>% 
  rename(name = race_4ce)
}

diagnosis_table1 <- summary_diagnosis %>% 
    bind_rows(
    sex_by_period %>%
      group_by(period) %>% 
      dplyr::summarise(concept_code = "Total", count = sum(count))
  ) %>%
  pivot_wider( names_from = period, values_from = count) %>% 
  mutate(diag = paste(concept_code, description, sep = ": ")) %>% 
  select(diag, early_pandemic, late_pandemic) %>% 
  rename(name = diag)
  
disorder_table1 <- summary_disorder_group_by_period %>% 
  bind_rows(
    dplyr::summarise(., disorder_group="Total", across(where(is.numeric), sum))
  ) %>% 
  mutate(result = format_counts(count, percentage)) %>% 
  select(-c(count, percentage)) %>% 
  pivot_wider( names_from = period, values_from = result ) %>% 
  replace_na( list(early_pandemic = "0 (0 %)", 
                   late_pandemic = "0 (0 %)")) %>% 
  rename(name = disorder_group)


hospit_table1 <- hospit_length %>% 
  mutate(median = glue::glue("{median} ({IQR})", median = length_hospt_median, IQR = IQR), 
         mean = glue::glue("{mean} ({sd})", mean = length_hospt_mean, sd = length_hospt_sd), 
         n_hospitalisations = as.character(n_hospitalisations)) %>% 
  select(median, mean, n_hospitalisations, period) %>% 
  pivot_longer(cols = c("n_hospitalisations", "median", "mean")) %>% 
  pivot_wider(names_from = period, values_from = value)

repeated_hospit_table1 <- repeated_hospitalisation %>% 
  pivot_longer(cols = c("non_repeated", "repeated")) %>% 
  pivot_wider(names_from = period, values_from = value)

if( raceAvailable == TRUE ){
  tests <- c("Sex" = sex_test, 
            "Race" = summary_race_test,
           "Diagnoses" = summary_diagnosis_test, 
           "Disorder groups" = summary_disorder_group_by_period_test, 
           "Hospitalisations length" = repeated_hospitalisation_test, 
           "Repeated hospitalisations" = repeated_hospitalisation_test) %>% 
    data.frame() %>% 
    round(2)
  names(tests) <- "pvalue"
  tests["Category"] <- row.names(tests)
  
  table1 <- list(
    "Sex" = sex_table1, 
    "Race" = race_table1,
    "Diagnoses" = diagnosis_table1,
    "Disorder groups" = disorder_table1, 
    "Hospitalisations length" = hospit_table1,
    "Repeated hospitalisations" = repeated_hospit_table1) %>% 
    lapply(function(df) mutate(df, across(.fns = as.character))) %>% 
    dplyr::bind_rows(.id = "Category") %>% 
    dplyr::left_join(tests, by = "Category")
  
}else{
  tests <- c("Sex" = sex_test, 
           "Diagnoses" = summary_diagnosis_test, 
           "Disorder groups" = summary_disorder_group_by_period_test, 
           "Hospitalisations length" = repeated_hospitalisation_test, 
           "Repeated hospitalisations" = repeated_hospitalisation_test) %>% 
  data.frame() %>% 
  round(2)
names(tests) <- "pvalue"
tests["Category"] <- row.names(tests)

table1 <- list(
  "Sex" = sex_table1, 
  "Diagnoses" = diagnosis_table1,
  "Disorder groups" = disorder_table1, 
  "Hospitalisations length" = hospit_table1,
  "Repeated hospitalisations" = repeated_hospit_table1) %>% 
  lapply(function(df) mutate(df, across(.fns = as.character))) %>% 
  dplyr::bind_rows(.id = "Category") %>% 
  dplyr::left_join(tests, by = "Category")
}

knitr::kable( table1 )
#library(kableExtra)
#table1 %>% 
#  kbl() %>%
#  kable_paper("striped", full_width = F) %>% 
#  column_spec(1, bold = T) %>%
#  collapse_rows(columns = c(1, 2, 5), valign = "top")

```

# Save the output in a text file. 
```{r, eval=FALSE}
```


# R session information
```{r}
sessionInfo()
```


##### To review 
## Temporal trend of hospitalized patients per day 

```{r, eval=FALSE }
# Creating encounter table with observation assigned to encounters
# 1. Creating encounters identifiers based on clinical_raw information about start and end date of an encounter --> n_hospitalisation
# 2. Assigning ICD codes to encounters (because they are assigned to the first or last day but usually are meaningful for the entire encounter)
encounters_obs <- clinical_raw %>% 
  filter(in_hospital == 1) %>% 
  select(patient_num, days_since_admission, n_hospitalisation) %>% 
  left_join(obs_raw[obs_raw$concept_type == "DIAG-ICD10",], on = c("patient_num", "days_since_admission")) %>% 
  drop_na(concept_type) %>% 
  distinct(patient_num, n_hospitalisation, concept_type, concept_code)
```

```{r, eval=FALSE}
obs_by_day = encounters_obs %>% 
  filter(concept_code %in% icdCodes$concept_code) %>% 
  left_join(clinical_raw[clinical_raw$in_hospital == 1, c("patient_num", "n_hospitalisation", "calendar_date")], 
            on = c("patient_num", "n_hospitalisation")) %>% 
  group_by(calendar_date) %>% 
  summarise(patient_count = n_distinct(patient_num)) %>% 
  mutate(calendar_date = as.Date(calendar_date, format=dateFormat), 
         period = if_else(calendar_date >= cut_off_date, "During", "Before")
  )

ggplot(obs_by_day, aes(x = calendar_date, y = patient_count, color = period))  +
  geom_line() +
  geom_vline(xintercept = as.Date(cut_off_date), linetype = "dashed") +
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "Hospitalized Patient Counts", 
       legend = "Period") +
  scale_color_brewer(palette = "Set1", name = "Period") 
ggsave("output/hospitalized_patients.png", width = 18, height = 9, units="cm")
```


## Hospitalized patients

```{r, eval=FALSE}
input_data %>% 
  group_by(date, period) %>% 
  summarise(hospitalized = n_distinct(patient_num))
```


